---
title: 'Lab 3: Reducing Crime (DRAFT)'
subtitle: 'W203 Statistics'
author: 'Luke Evans and Daniel Rasband'
date: \today
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
 - Motivation: To address the causes of crime, and/or to promote the crime inhibitors
 - Enable the political campaign to develop policies that will address crime in North Carolina
 - Identification of key focus areas that: Address causes of crime; capitalize on crime inhibitors
 - Not going to include costing of any measures
 - Exclude anything else?

## Exploratory Data Analysis
  - 2 types of factor: crime drivers; crime surpressors 
  - Some exceptional data in county 185, presents weekly service wages 10 times larger than any other weekly wage which is skewing the dataset
  - Some exceptional data in county 115 where police per capita is double the nearest value
  - Crime drivers apprear to be:
    - density: Population Density
    - taxpc: Tax revenue per capita
    - pctmin80: percent of minority in the population
    - pctymle: percent of young male in the population 
    - Most of the wages - as wage rises, crime rates increase. Fed employees drive this most, whereas state employees drive this least
  - Crime inhibitors appear to be:
    - prbarr: The probability of arrest
    - prbconv: The probability of conviction
    
    Note probability of a custodial term nor the length of that term does not appear to be a strong crime deterant
    Police per capita is an odd one, probably good one to remove?
  
```{r}
# Load the data from CSV.
crime.dat <- read.csv('crime_v2.csv')
library(car)
scatterplotMatrix( ~  + crmrte + density + taxpc + pctmin80 + pctymle, data = crime.dat,
                   main = "Scatterplot Matrix for Key crime drivers")
scatterplotMatrix( ~  + crmrte + prbarr + prbconv, data = crime.dat,
                   main = "Scatterplot Matrix for Key crime inhibitors")
```

  

## Data Integrity and Cleaning

```{r}
# Clean the data: Rows 92-97 are empty, so removing them.
crime.dat <- na.omit(crime.dat)
```

## The Model

## The Regression Table

## Omitted Variables
 - Do we need all the wages, if we have a value for tax revenue per capita?
 - police per capita?
## Conclusion